---
title: "Instrumental Variables (IV) Estimation"
format: 
  html:
    toc: true
    css: [webex.css]
    include-after-body: [webex.js]
    embed-resources: true 
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(webexercises)
```

```{r, echo = FALSE, results='asis'}
# Uncomment to change widget colours:
#style_widgets(incorrect = "goldenrod", correct = "purple", highlight = "firebrick")
```


## Introduction 

In the Panel-data section we discussed an example of using information from data on twins to estimate the effect of schooling on wages. We proposed the following model

$$
\log y_{fi} = \beta_0 + \beta_1 s_{fi} + \mathbf{x}'_f
\mathbf{b} + a_f + u_{fi} \qquad i=1,2
$$

Here, subscript $f$ stands for family, while subscript $i = 1,2$ indicated twin 1 or 2 in family $f$. The reason why we suspected that there may be issues in estimating the above model, either by pooled OLS, and even more so when using a difference model (calculating the difference between siblings) was that there is the distinct possibility of individuals misreporting their own amount of schooling $s_{fi}$.

As discussed in the endogeneity section, this will cause the $s_{it}$ variable to be endogenous, causing a bias in the estimation of $\beta_1$.

The method introduced here, IV Estimation, is a tool that, sometimes, can be used to deal with this issue. Towards the end of the chapter we will return to this example.








## Some econometric details

Both the long-levels and short-differenced regressions can be written as

$$\log y = \beta_0 + \beta_1 s^* + u,$$

where both $s^*$, the self-report, and $z^*$, the cross-sibling report,
are measured with error:
\[
s = s^* + e \qquad z = z^* + m.
\]
We know from Section~\ref{sub:ME} that the (asymptotic) bias in OLS is
given by:
\[
\text{plim}\betahat_1 = \beta_1 \left(1-\frac{\Var(e)}{\Var(s)} \right).
\]
Now the IV estimator is given by:
\[
\betatilde_1 = \frac{\Covhat(\log y,z)}{\Covhat(s,z)} =
\frac{\Covhat(\beta_0 + \beta_1 s^* + u,z)}{\Covhat(s^* + e,z)}.
\]

First consider the asymptotic bias in IV when the cross-sibling report
is not measured with error, $m=0$. Whilst not the case here, it shows
how IV can be used to solve ME problems in general.
\begin{multline*}
\text{plim}(\betatilde_1) =
\frac{\text{plim}[\Covhat(\beta_0 + \beta_1 s^* + u,z)]}{\text{plim}[\Covhat(s^* + e,z)]}=
\frac{\Cov (\beta_0 + \beta_1 s^* + u,z) } {\Cov (s^* + e,z) } =  \\
\frac{\beta_1 \Cov (s^*,z) + \Cov(u,z)}{\Cov (s^* + e,z) } =
\frac{\beta_1 \Cov (s^*,z) + 0}{\Cov (s^*,z) } = \beta_1.
\end{multline*}
The key assumption, in general, is that the instrument is uncorrelated
with the measurement error,  $\Cov (e,z)=0$ (as well as the equation
error $u$).

Now consider the asymptotic bias in IV when the cross-sibling report
\ul{is} measured with error, $m \neq 0$.
\begin{multline*}
\text{plim}(\betatilde_1) =
\frac{\text{plim}[\Covhat(\beta_0 + \beta_1 s^* + u,z)]}{\text{plim}[\Covhat(s^* + e,z)]}=
\frac{\Cov (\beta_0 + \beta_1 s^* + u,z^* + m) } {\Cov (s^* + e,z^* + m) } =  \\
\frac{\beta_1 \Cov (s^*,z^* + m) + \Cov(u,z^* + m)}{\Cov (s^* + e,z^* + m) } =
\frac{\beta_1 \Cov (s^*,z^* + m) + 0}{\Cov (s^* + e,z^* + m) } = \dots?
\end{multline*}

For this to equal $\beta_1$, we need:
\[
\Cov (s^*, m) = \Cov (z^*, e)  =\Cov (e, m) =0.
\]
There is no reason why the ME in cross-sibling report, $m$, should be
correlated with the true own years-of-education.  And the same the other
way round.  The correlation between the two MEs is more contentious.  MM
say:  ``Uncorrelatedness of reporting errors across siblings is a strong
assumption, but is a natural starting point \dots''

If $\Cov (e, m) \neq 0$, then $z$ is not an exogenous instrument.
However, exludability is a very reasonable assumption, and the $F$ tests
for relevance are easily big enough (observed $s$ is strongly correlated
with observed $z$, even in differences).






## Reading

Reading

* Angrist, J.D. and J-S. Pischke (2015) Mastering Metrics. The Path from Cause to Effect. Chp 3, Princeton University Press: Princeton.

* Cunningham, S. (2026), Chapter 7 Instrumental Variables, [https://mixtape.scunning.com/09-difference_in_differences](https://mixtape.scunning.com/09-difference_in_differences) (accessed 17/3/2026)

* Wooldridge, J. (2025) Introductory Econometrics: A Modern Approach, Cengage, 8th edition, Section 15.1-15.4. Policy Analysis with Pooled Cross-Section Data.




# DELETE

## Example Questions

### Fill-In-The-Blanks (`fitb()`)

Create fill-in-the-blank questions using `fitb()`, providing the answer as the first argument.

- 2 + 2 is `r fitb(4)`

You can also create these questions dynamically, using variables from your R session.

```{r}
x <- sample(2:8, 1)
```

- The square root of `r x^2` is: `r fitb(x)`

The blanks are case-sensitive; if you don't care about case, use the argument `ignore_case = TRUE`.

- What is the letter after D? `r fitb("E", ignore_case = TRUE)`

If you want to ignore differences in whitespace use, use the argument `ignore_ws = TRUE` (which is the default) and include spaces in your answer anywhere they could be acceptable.

- How do you load the tidyverse package? `r fitb(c("library( tidyverse )", "library( \"tidyverse\" )", "library( 'tidyverse' )"), ignore_ws = TRUE, width = "20")`

You can set more than one possible correct answer by setting the answers as a vector.

- Type a vowel: `r fitb(c("A", "E", "I", "O" , "U"), ignore_case = TRUE)`

You can use regular expressions to test answers against more complex rules.

- Type any 3 letters: `r fitb("^[a-zA-Z]{3}$", width = 3, regex = TRUE)`

### Multiple Choice (`mcq()`)

- "Never gonna give you up, never gonna: `r mcq(c("let you go", "turn you down", "run away", answer = "let you down"))`"
- "I `r mcq(c(answer = "bless the rains", "guess it rains", "sense the rain"))` down in Africa" -Toto

### True or False (`torf()`)

- True or False? You can permute values in a vector using `sample()`. `r torf(TRUE)`

### Longer MCQs (`longmcq()`)

When your answers are very long, sometimes a drop-down select box gets formatted oddly. You can use `longmcq()` to deal with this. Since the answers are long, It's probably best to set up the options inside an R chunk with `echo=FALSE`. 

**What is a p-value?**

```{r}
opts_p <- c(
   "the probability that the null hypothesis is true",
   answer = "the probability of the observed, or more extreme, data, under the assumption that the null-hypothesis is true",
   "the probability of making an error in your conclusion"
)
```

`r longmcq(opts_p)`

**What is true about a 95% confidence interval of the mean?**

```{r}
# use sample() to randomise the order
opts_ci <- sample(c(
  answer = "if you repeated the process many times, 95% of intervals calculated in this way contain the true mean",
  "there is a 95% probability that the true mean lies within this range",
  "95% of the data fall within this range"
))
```

`r longmcq(opts_ci)`

## Checked sections

Create sections with the class `webex-check` to add a button that hides feedback until it is pressed. Add the class `webex-box` to draw a box around the section (or use your own styles).

::: {.webex-check .webex-box}

I am going to learn a lot: `r torf(TRUE)`

```{r, results='asis'}
opts <- c(
   "the probability that the null hypothesis is true",
   answer = "the probability of the observed, or more extreme, data, under the assumption that the null-hypothesis is true",
   "the probability of making an error in your conclusion"
)

cat("What is a p-value?", longmcq(opts))
```

:::



## Hidden solutions and hints

You can fence off a solution area that will be hidden behind a button using `hide()` before the solution and `unhide()` after, each as inline R code.  Pass the text you want to appear on the button to the `hide()` function.

If the solution is a code chunk, instead of using `hide()` and `unhide()`, simply set the `webex.hide` chunk option to TRUE, or set it to the string you wish to display on the button.

**Recreate the scatterplot below, using the built-in `cars` dataset.**

```{r}
with(cars, plot(speed, dist))
```


`r hide("I need a hint")`

See the documentation for `plot()` (`?plot`)

`r unhide()`

<!-- note: you could also just set webex.hide to TRUE -->

```{r, echo = TRUE, eval = FALSE}
#| webex.hide: "Click here to see the solution"
plot(cars$speed, cars$dist)
```
