---
title: "Introduction to Panel Data Models"
output:
 html_document:
   df_print: paged
   keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning = FALSE)
```


# Preparing your workfile

We add the basic libraries needed for this week's work:

```{r}
library(tidyverse)    # for almost all data handling tasks
library(ggplot2)      # to produce nice graphics
library(stargazer)    # to produce nice results tables
library(haven)        # to import stata file
library(AER)          # access to HS robust standard errors
source("stargazer_HC.r")  # includes the robust regression display
```

As we are using panel methods we also require an additional package `plm`.

```{r}
#install.packages("plm") # only execute this if plm is not installed yet
library(plm)
```


# Introduction

The data are an extract from the [Understanding Society Survey](https://www.understandingsociety.ac.uk/) (formerly the British Household Survey Panel).

# Data Upload - and understanding data structure

Upload the data, which are saved in a STATA datafile (extension `.dta`). There is a function which loads STATA file. It is called `read_dta` and is supplied by the `haven` package.

```{r}
data_USoc <- read_dta("../data/20222_USoc_extract.dta")
data_USoc <- as.data.frame(data_USoc)    # ensure data frame structure
names(data_USoc)
```

Let us ensure that categorical variables are stored as `factor` variables. It is easiest to work with these in R.

```{r}
data_USoc$region <- as_factor(data_USoc$region)
data_USoc$male <- as_factor(data_USoc$male)
data_USoc$degree <- as_factor(data_USoc$degree)
data_USoc$race <- as_factor(data_USoc$race)
```

The pay information (`paygu`) is provided as a measure of the (usual) gross pay per month. As workers work for varying numbers of hours per week (`jbhrs`) we divide the monthly pay by the approximate monthly hours (`4*jbhrs`). We shall also adjust for increasing price levels (as measured by `cpi`). These two adjustments leave us with an inflation adjusted hourly wage. We call this variable `hrpay` and also calculate the natural log of this variable (`lnhrpay`).

```{r}
data_USoc <- data_USoc %>%
              mutate(hrpay = paygu/(jbhrs*4)/(cpi/100)) %>%
              mutate(lnhrpay = log(hrpay))
```

As we wanted to save these additional variables we assign the result of the operation to `data_USoc`.

We will also use the logarithm of the unemployment rate

```{r}
data_USoc <- data_USoc %>%
              mutate(lnurate=log(urate)) 
```

# Understanding the Panel Structure

To explain the meaning of these let us just pick out all the observations that pertain to one particular individual (`pidp == 272395767`). The following command does the following in words: "Take `data_USoc` filter/keep all observations which belong to individual pidp == 272395767, then select a list of variables (we don't need to see all 14 variables) and print the result":

```{r}
data_USoc %>% filter(pidp == 272395767) %>% 
              select(c("pidp","male","wave","year","paygu","age","educ")) %>% 
              print()
```

The same person (female) was observed three years in a row (from 2009 to 2011). Their gross monthly income changed, as did, of course, their age, but not their education. This particular person was observed in three consequitive waves. Let's se whether this is a commom pattern.

In the context of this exercise we will ignore the second wave and only look at waves 1 and 3.

```{r}
data_USoc <- data_USoc %>%  
  filter(wave != 2) %>% 
  filter(!is.na(lnhrpay))
```

The code below figures how many waves we have for each individual (1 or 2) and then saves this in a new variable (`n_wave). This information will be used later as we may want to know whether only using observations for which we have both waves makes a difference to the analysis.

```{r}
data_USoc <- data_USoc %>% 
                group_by(pidp) %>% 
                mutate(n_wave = n()) 
```

Now we need to let R know that we are dealing with panel data. This is why we loaded up the `plm` library which contains the `plm.data` function. Using the `index = c("pidp","wave")` we let the function know what identifies the individuals and what identifies the wave.

```{r}
pdata_USoc <- pdata.frame(data_USoc, index = c("pidp","wave")) # defines the panel dimensions
```

We saved the output in `pdata_USoc` and we will use this for any panel data estimations.

When dealing with panel data it is super useful to understand in how many and in which wave individuals are prepresented. We already calculated the `n_wave` variable which tells us in how many of our remaining two waves observations are represented. We also have information (`wave`) on which wave someone is represented in. To disentagle this we merely need a contingency `table` of the `n_wave` and `waves` variables.

```{r}
table(pdata_USoc$n_wave,pdata_USoc$wave, dnn = c("n_waves","waves"))
```

Naturally the 13078 respondents which have two observations (`n_wave == 2)` are represented in waves 1 and 3. Then we have (`n_wave == 1`) 9666 respondents which are represented in wave 1 only and 4112 which are represented in wave 3 only.

For the respondents for which we have 2 waves of observations we can actually calculate a difference, or change in variables. This will become important in a later model estimation (although for that we could let R do the work in the background) and hence we will calculate these variables explicitly, here for `lnhrpay` and `lnurate`.

```{r}
# the lag function below will recognise the panel nature of the data and 
# will only calculate lags for individuals
# we also need to specify that we are calculating k=2 step lag as
# we calculate the difference between wave 3 and 1

Dlnhrpay <- pdata_USoc$lnhrpay-lag(pdata_USoc$lnhrpay,k=2)
Dlnurate <- pdata_USoc$lnurate-lag(pdata_USoc$lnurate,k=2)

pdata_USoc$Dlnhrpay <- Dlnhrpay   # add the new series to the dataframe
pdata_USoc$Dlnurate <- Dlnurate
```

For a later purpose we will also identify all individuals who moved from one region to another between waves 1 and 3. It is not so important to understand this code.

```{r}
temp <- pdata_USoc   # create a temporary dataframe

temp <- temp %>%  filter(n_wave == 2) %>%   # only keep individuals with two waves
                  group_by(pidp) %>%        # group data by individual
                  mutate(move = ifelse(length(unique(region))==1,"no move","move")) %>% 
                  select(pidp,wave,move)    # only keep these 3 variables

# the move variable will take the value "no move" if both regions are identical  
# and "move" if there are two different regions  

temp$move <- as_factor(temp$move)  # convert to factor variable

# the following merges the new variable into the pdata_USoc dataframe
pdata_USoc <- merge(pdata_USoc,temp,all.x = TRUE) 
```

Let's check how many movers there are.

```{r}
pdata_USoc %>%  count(move)
```

So ther are 352 observations associated with movers. That means that there are 176 movers.

# Some data descriptions

We will use the `lnhrpay` and the `urate` variables below. We therefore will have a look at these variables.

```{r}
stargazer(pdata_USoc[,c("lnhrpay","urate","year","Dlnhrpay","Dlnurate")],type = "text")
```

Let us look at some summary statistics grouped by region

```{r}
pdata_USoc %>% group_by(region) %>% 
              summarise(n = n(), mean_lnhrpay = mean(lnhrpay),mean_urate = mean(urate))

```

Below we will want to use the mean `lnhrpay` and mean `lnurate` as calculated for every region-year. The following will group the data by region-wave (as we have 12 regions and 2 waves we will 24 such groups). This is similar to the above command but note that we start with `pdata_USoc <-` to ensure that the calculated average wage and unemployment rate values are added as variables to the data frame. Also, instead of `summarise` (which displays the calculated statistics) we use the `mutate` function as we want the calculated series to be saved in the data frame.

```{r}
pdata_USoc <- pdata_USoc %>% 
              group_by(region,year) %>% 
              mutate(mean_lnhrpay = mean(lnhrpay),mean_urate = mean(urate)) 
```

This will retain the number of observations but will now have added the relevant region-year means of `lnhrpay` and `urate` to each observation.

# Estimating Models

## Pooled Regression

We start by estimating a model which does not use the panel nature of the data.

```{r}
POLS0 <- lm(lnhrpay~lnurate, data = pdata_USoc)
stargazer_HC(POLS0)
```

Let's add the predicted model values to the data frame. As our explanatory variable only has 24 different values (12 regions and two years) we will only get 24 different predicted values.

```{r}
pdata_USoc$pred_POLS0 <- POLS0$fitted.values
```

Here we basically used all observations available, whether they were from wave 1 or 3. We **pooled** the observations and hence we could use our normal `lm` function to estimate this model. The `plm` package we imported earlier has a few panel specific tricks up its sleeve and we could estimate this model with the `plm` function.

```{r}
POLS0a <- plm(lnhrpay~lnurate, data = pdata_USoc, model = "pooling")
stargazer_HC(POLS0a)
```

Now we plot the predicted values and compare them against the region-year's average values (`mean_lnhrpay`).

```{r}
# pdf("Lecture6plot_Pooled.pdf",width = 5.5, height = 4) # uncomment to save as pdf
ggplot(pdata_USoc, aes(x=lnurate,y=pred_POLS0)) +
  geom_point(aes(colour = "red")) +
  geom_line(aes(colour = "red")) +
  geom_point(aes(y = mean_lnhrpay,colour = "blue")) +
  ggtitle("Predicted values - Pooled OLS") +
  ylab("Log Hourly Wage") +
  xlab("Log Unemployment Rate") +
  scale_colour_manual(name="Data/Model", values = c(red = "red", blue = "blue"),
                      labels=c("Data", "POLS"))
# dev.off() # uncomment to save as pdf
```

Now we will include a dummy variable for `wave == 3`. The `wave` variable is a factor variable with two levels (1 and 3) for waves 1 and 3.

```{r}
POLS1 <- lm(lnhrpay~lnurate+wave, data = pdata_USoc)
stargazer_HC(POLS1)
```

The first wave is the base category of `wave` and hence is not included. Alternatively this could be estimated using the `plm` package

```{r}
POLS1a <- plm(lnhrpay~lnurate+wave, data = pdata_USoc, model = "pooling")
stargazer_HC(POLS1a)
```

This regression will have observations for individuals for which we only observe one wave (`n_wave == 1`). Let's restrict the analysis to only individuals for which we have two waves (`n_wave == 2`).

```{r}
POLS2 <- lm(lnhrpay~lnurate+wave, data = pdata_USoc, subset = (n_wave ==2))
stargazer_HC(POLS2)
```

or using the `plm` function

```{r}
POLS2a <- plm(lnhrpay~lnurate+wave, data = pdata_USoc, subset = (n_wave ==2), model = "pooling")
stargazer_HC(POLS2a)
```

## First Difference Estimation

Now we estimate a first difference (FD) model. We will only do this using the `plm` function. If we were to use the `lm` function we had to first calculate differenced series (which we have done on this occasion, but only to illustrate the mechanics). Before we estimate the model let's look at the data for a few respondents.

```{r}
pdata_USoc %>% filter(pidp %in% c("3915445","68001367","68004087","68195851")) %>% 
              select(c("pidp","male","wave","lnhrpay","Dlnhrpay","lnurate", "Dlnurate")) %>% 
              print()
```

When estimating a FD model we are basically running a regression of `Dlnhrpay` on `Dlnurate`. Respondents for whom we do not have two waves will not be used in such a model. The calculation of the `Dlnhrpay` and `Dlnurate` series happens automatically inside the `plm` function when we specify `model = "fd"`.

```{r}
FD1a <- plm(lnhrpay~lnurate, data = pdata_USoc, subset = (n_wave ==2), model = "fd")
stargazer_HC(FD1a)
```

Note the number of observations which corresponds to the number of individuals for which we had observations from both waves. We can show a scatter plot of the available difference observations and the regression line estimated by `FD1a`.

```{r}
# pdf("Lecture6plot_FD_R.pdf",width = 5.5, height = 4) # uncomment to save as pdf
ggplot(pdata_USoc, aes(x=Dlnurate,y=Dlnhrpay,color=move)) +
  geom_point() +
  geom_abline(intercept = FD1a$coefficients[1], slope = FD1a$coefficients[2],colour = "blue") +
  ggtitle("Scatter plot and FD regression line") +
  ylab("D(Log Hourly Wage)") +
  xlab("D(Log Unemployment Rate)")
# dev.off() # uncomment to save as pdf
```

As you can see, there is no obvious relationship between the changes in hourly pay and the respective local unemployment rate.

Now we will show models, `POLS0a`, `POLS1a`, `POLS2a` and `FD1a` in one table. In previous tables you may have seen that the F-stat takes up a lot of space and hence we use the `omit_stat` option to indicate that we do not want to see the F-statistic.

```{r}
stargazer_HC(POLS0a,POLS1a,POLS2a,FD1a,omit.stat = "f")
```

## Fixed Effects estimation

What we have basically done by estimating the model as a FD model is to eliminate the "between" variation, estimating the effect size $\beta_1$ only on the basis of the "within" variation.

The same can be achieved by including individual level dummy variables. To illustrate this we will create a little test dataset with many fewer observations. In particular we pick males with race declared as "other" and for whom we have two observations. The only reason we chose this particular subgroup is to get a small set of individuals. 

```{r}
pdata_USoc_test <- pdata_USoc %>% filter(n_wave ==2) %>% 
                      filter(male == "male", race == "other")
```

This results in 160 observations for 80 individuals. Let us first re-estimate the FD model from earlier. The "-1" in the specification removes the constant as that gets differenced out as well. We will only show the estimated coefficient to $lnurate$.

```{r}
FD1a_test <- plm(lnhrpay~lnurate-1, data = pdata_USoc_test, model = "fd")
stargazer_HC(FD1a_test, keep = "lnurate")
```

The actual value is of course different to the above estimated value and by itself is not of importance.

Now we include dummy variables for each individual (`+pidp`). As `pidp` is a factor variable this will automatically include a dummy variable for all but one of the individuals.

```{r}
FE1a_test <- plm(lnhrpay~lnurate+pidp, data = pdata_USoc_test, model = "pooling")
```

```{r}
stargazer_HC(FD1a_test, FE1a_test, keep = "lnurate")
```

You can see that the estimated coefficients are identical. There are all sorts of differences between the two models, like the number of observations, the $R^2$ and the standard errors. We shall not discuss these issues further but note that in terms of estimating the effect coefficient $\beta_1$ the result is identical. The coefficient has only been estimated using "within" variation in the data. 

The reason that we reduced the dataset to demonstrate this was that otherwise there were more than 13,000 individuals which would require the inclusion of that many dummy variables. That is computationally very demanding and should be avoided.

In fact there is yet an additional alternative way of achieving the same, i.e. estimating the effect on the basis of the "within" variation only. The `plm` package offers the option `model = "within"`:

```{r}
FE1b_test <- plm(lnhrpay~lnurate, data = pdata_USoc_test, model = "within")
stargazer_HC(FD1a_test, FE1a_test, FE1b_test, keep = "lnurate")
```

Again you see that the effect size is identical. What this does is to deduct the average for each individual from each variable before using them in the regression: $y_{it} - \bar{y_i}$ and $x_{it} - \bar{x_i}$ where $\bar{x_i}$ is the average of all $x$ observations for individual $i$. 