---
title: "Panel Data Methods"
format: 
  html:
    toc : true
    css: [webex.css]
    include-after-body: [webex.js]
    embed-resources: true 
editor: 
  render-on-save: true
  markdown: 
    wrap: 72
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(webexercises)
```

```{r, echo = FALSE, results='asis'}
# Uncomment to change widget colours:
#style_widgets(incorrect = "goldenrod", correct = "purple", highlight = "firebrick")
```

# Introduction to Panel Data Structures

We shall begin by introducing Panel Data structures. Being clear about
data structures is a prerequisite for successful empirical work.

Previously we discussed that parameter estimator bias can be caused by
omitted variables sometimes also called unobserved heterogeneity. There
will be situations when using panel data will give us a route to
removing this bias.

## Two-period panel data

Suppose we have a dataset comprising $N$ individuals (or other micro
unit) observed exactly $T=2$ times.

| $i$ | $t$ | $y_{it}$ | $1$ | $d^1_t$ | $d^2_t$ | $x_{it}$ | $a_{i}$ | $u_{it}$ |
|-----|-----|----------|-----|---------|---------|----------|---------|----------|
| 1   | 1   | $y_{11}$ | 1   | 1       | 0       | $x_{11}$ | $a_{1}$ | $u_{11}$ |
| 1   | 2   | $y_{12}$ | 1   | 0       | 1       | $x_{12}$ | $a_{1}$ | $u_{12}$ |
| 2   | 1   | $y_{21}$ | 1   | 1       | 0       | $x_{21}$ | $a_{2}$ | $u_{21}$ |
| 2   | 2   | $y_{22}$ | 1   | 0       | 1       | $x_{22}$ | $a_{2}$ | $u_{22}$ |
| ... | ... | ...      | ... | ...     | ...     | ...      | ...     | ...      |
| $N$ | 1   | $y_{N1}$ | 1   | 1       | 0       | $x_{N1}$ | $a_{N}$ | $u_{N1}$ |
| $N$ | 2   | $y_{N2}$ | 1   | 0       | 1       | $x_{N2}$ | $a_{N}$ | $u_{N2}$ |

The variables we observe are

-   $y_{it}$ and $x_{it}$; both vary over both $i$ and $t$.
-   $d^1_t$ is a period 1 time dummy; it only varies over $t$.
-   $d^2_t$ is a period\~2 time dummy; it only varies over $t$.
-   $1$ is the constant.

whereas the variables not observed are:

-   $a_i$ does not vary over $t$. It is a so-called **fixed effect**, or
    **unobserved heterogeneity**. Think about a personal characteristic
    of subject $i$ that is constant through time. This could be things
    like general levels of ability or intelligence.
-   $u_{it}$ (not shown) is an **idiosyncratic error**. It is
    uncorrelated with everything (like most error terms). In particular:

$$
Cov(x_{i1},u_{i1})=Cov(x_{i2},u_{i2})=0 ~~~(AFD1)
$$

Let's work with the following example:

-   $y_{it}$ is log earnings (`lhrpay`)
-   $x_{it}$ is years of schooling (`educ`)
-   $a_{i}$ is unobserved innate ability (`abil`)

If you wanted to establish whether years of schooling had some statistically significant impact on earnings you will be tempted to estimate the following  wage-education model (essentially the same as in the Omitted Variable Bias Section):

$$
y_{it} = \beta_0 + \delta_0 d^2_t + \beta_1 x_{it} + \beta_2 a_{i} + u_{it}.
$$ {#eq-woods}

Note that we included a dummy variable for the 2nd period, \$d\^2_t \$.
Because of the dummy-variable trap, one of two time dummies must be
dropped, here $d^1_t$. (Recall one cannot have a constant, male dummy
**and** a female dummy in the same regression and the same is true for time dummies. If there are two periods only one time dummy can be included).

The crucial problem occurs when realising that $a_i$ is unobservable and therefore will have to be omitted. If the omitted variable $a_i$ is correlated
with $x_{it}$ then we are basically dealing with the case of an omitted variable which causes an Omitted
Variable Bias if $a_i$ is correlated with $X_{it}$.

Here, ability is correlated with education. If $a_i$ were **not**
correlated with $x_{it}$, the OVB formula says that OLS is unbiased and we
wouldn't have a problem in the first place. The way that panel data can
help is through the crucial assumption that $a_i$ is constant over time.
This is an untestable assumption as $a_i$ is not observed.

The above equation is also called a **fixed effects** model.

# The First Difference estimator (FD) {#sec-FD}

The First Difference estimator makes use of this setup to estimate
$\beta_1$.

## Differencing out $a_i$

We could estimate @eq-woods by OLS, treating $a_i + u_{it}$ as a single
**composite error**. This estimator is called **Pooled OLS** (POLS) as it uses all data pooled across both time periods, hence $2N$ observations. However, we have already assumed that $Cov(x_{it},a_i)\neq 0$ which
means that Pooled OLS is biased (and inconsistent). This is often called
**heterogeneity bias** which is just a different name for OVB in this circumstance.

**However**, with panel data, we can difference-out the unobserved fixed
effect $a_i$. Let's write @eq-woods for both time periods $t=1$ and
$t=2$.

$$
\begin{gather*}
y_{i1} = \beta_0            + \beta_1 x_{i1}  + a_i + u_{i1} \quad (t=1)\\
y_{i2} = \beta_0 + \delta_0 + \beta_1 x_{i2}  + a_i + u_{i2} \quad (t=2)
\end{gather*}
$$

Subtracting the $t=1$ equation from the $t=2$ equation gives

$$
\begin{equation}
  \Delta y_{i} = \beta_1 \Delta x_{i} + \Delta u_{i}.
\end{equation}
$$ {#eq-mee}

Importantly, $\beta_1$, the coefficient to $\Delta x_{i}$, is exactly the same coefficient as the coefficient to $x_{it}$ in the original regression. Importantly, the fixed effect $a_i$ has been differenced out because it is (assumed to be) the same in both periods. This means that this term will not end up in an error term where it could cause trouble through any correlation with an explanatory variable.

To see what is happening, look at the schematic data above, and subtract
the first row from the second for each $i$:

| $i$ | $t$ | $y_{it}$ | 1 | $d^1_{t}$ | $d^2_{t}$ | $x_{it}$ | $\Delta y_{it}$ | $\Delta 1$ | $\Delta d^1_{t}$ | $\Delta d^2_{t}$ | $\Delta x_{it}$ |
|------|------|------|------|------|------|------|------|------|------|------|------|
| 1 | 1 | $y_{11}$ | 1 | 1 | 0 | $x_{11}$ | . | . | . | . | . |
| 1 | 2 | $y_{12}$ | 1 | 0 | 1 | $x_{12}$ | $\Delta y_{1}$ | 0 | --1 | 1 | $\Delta x_{1}$ |
| 2 | 1 | $y_{21}$ | 1 | 1 | 0 | $x_{21}$ | . | . | . | . . | . |
| 2 | 2 | $y_{22}$ | 1 | 0 | 1 | $x_{22}$ | $\Delta y_{2}$ | 0 | --1 | 1 | $\Delta x_{2}$ |
| ... | ... | ... | ... | ... | ... | ... | ... | ... | ... | ... | ... |
| $N$ | 1 | $y_{N1}$ | 1 | 1 | 0 | $x_{N1}$ | . | . | . | . | . |
| $N$ | 2 | $y_{N2}$ | 1 | 0 | 1 | $x_{N2}$ | $\Delta y_{N}$ | 0 | --1 | 1 | $\Delta x_{N}$ |

Expressing @eq-mee in words, the change in $y_{it}$ between periods 1
and 2 is regressed on the change in the regressor(s) and a constant,
using $N$ obs. Note that the variables are not defined for period 1 and
so the dataset comprises just $N$ obs, whereas regressing @eq-woods
could be estimated on $2N$ observations.

Just to be clear about the variables that are being used in @eq-mee:

-   $\Delta y_{i} \equiv y_{i2}-y_{i1}$ is the \textbf{difference} of
    $y_{i}$.
-   $\Delta x_{i} \equiv x_{i2}-x_{i1}$ is the difference in the explanatory variable. (In a multiple regression all the regressors are differenced.)
-   $\beta_0$ and $\beta_0+\delta_0$ are "macro" effects in that they
    affect all individuals identically, in the same time period. But
    only $\delta_0$ is identified. $\beta_0$ drops out because
    $\Delta 1 =0 = \Delta d^1_{t} +\Delta d^2_{t}$.
-   $\delta_0$ is the parameter on $\Delta d^2_t$, which is the new
    regression constant.

When OLS is used on this transformed model, it is labelled the
**first-difference estimator** (FD). In this example, it is a Simple
Regression. In general, FD is a multiple regression where **all**
variables are differenced.

The important issue is whether the assumptions for OLS are satisfied? If
yes, FD is unbiased. Recall that the assumptions are:

-   The RHS variable is exogenous (uncorrelated with the error);
-   The RHS variable cannot be constant.

## Exogeneity

The first assumption asks whether $\Delta x_{i}$ is uncorrelated with
$\Delta u_{i}$?

$$
\begin{align*}
Cov (\Delta x_i, \Delta u_i) &=
Cov [(x_{i2}-x_{i1}),(u_{i2}-u_{i1})] \\
& =Cov(x_{i2},u_{i2}) +
Cov(x_{i1},u_{i1}) -
Cov(x_{i1},u_{i2}) -
Cov(x_{i2},u_{i1}) \\ 
&=0.
\end{align*}
$$

All 4 covariances need to be zero. We have already assumed the first
pair are zero (AFD1). If AFD1 is true, there is no reason why the second
pair ("across" $t=1,2$) should not be zero either.

To summarise, if $x$ and $u$ are uncorrelated, then so are $\Delta x$
and $\Delta u$. But this is an assumption. One needs to
assume that there is no unobserved variation in the error terms that is
correlated to the $x$ variable. What has been gained through the
first-difference setup is that any such variation in the error term that
is constant through time by $i$ has been removed. We do not have to
worry any longer about such unobserved variation.

::: callout-note
### Practise Question

Consider the setup where

-   $y_{it}$ is log earnings (`lhrpay`) of individual $i$ in period $t$.
-   $x_{it}$ is a variable encoding how many hours individual $i$ spent
    on learning new skills in period $t$ Which of the following
    potentially unobserved heterogeneity will be removed/taken care of
    by applying a FD estimation in a context where you have a panel
    dataset for many individuals $i$ and two time periods for each
    individual.

```{r}
#| echo: false
opts_p <- c(
   answer = "different general ability by individual $i$",
   "different levels of economic growth in individual $i$'s area",
   answer = "different personalities of individuals $i$",
   "different life circumstances for individuals $i$ at different $t$"
)

```

`r longmcq(opts_p)`

`r hide("Can you help me understand the answer")` It is important to
understand that only unobserved variation that remains constant for
individual $i$ across the time periods is differenced out. This is why
economic growth (which varies through time) and life circumstances which
often change from one year to the next, are not controlled for by using
a FD estimation. `r unhide()`
:::

## Variation in $\Delta x_i$

The second assumption says that each $\Delta x_{i}$ must have some
variation across $i$. Age, gender, and (mostly) education are good examples of
$x_{i}$ variables that are constant through time. To see what is happening, observe what
happens when we first-difference the male dummy $m$, age $a$ and
education $e$:

| $i$ | $t$ | $y_{it}$ | $d^2_t$ | $m$ | $a$ | $e$ |   | $\Delta y_{it}$ | $\Delta d^2_{t}$ | $\Delta m$ | $\Delta a$ | $\Delta e$? |
|------|------|------|------|------|------|------|------|------|------|------|------|------|
| 1 | 1 | $y_{11}$ | 0 | 1 | 18 | 11 |  | . | . | . | . |  |
| 1 | 2 | $y_{12}$ | 1 | 1 | 24 | 11 |  | $\Delta y_{1}$ | 1 | 0 | 6 |  |
| 2 | 1 | $y_{21}$ | 0 | 0 | 28 | 13 |  | . | . . | . | . |  |
| 2 | 2 | $y_{22}$ | 1 | 0 | 34 | 13 |  | $\Delta y_{2}$ | 1 | 0 | 6 |  |
| ... | ... | ... | ... | ... | ... | ... |  | ... | ... | ... | ... |  |
| $N$ | 1 | $y_{N1}$ | 0 | 1 | 51 | 16 |  | . | . | . | . |  |
| $N$ | 2 | $y_{N2}$ | 1 | 1 | 57 | 16 |  | $\Delta y_{N}$ | 1 | 0 | 6 |  |


::: callout-note
### Practise Question

What values should be in the $\Delta e$ column?

```{r}
#| echo: false
opts_p <- c(
   answer = "0",
   "1",
   "2",
   "-1",
   "various values"
)

```

`r longmcq(opts_p)`

`r hide("Can you help me understand the answer")` 
Certainly for the data shown in the table the years of education do not change and therefore it is 0 for all observations. This is quite typical for datasets which only contains people in working age. Although this would not be true any longer if some people in the dataset would actually go back to school or university.
`r unhide()`
:::

This is important, because some issues (education, discrimination etc)
cannot be investigated using this approach (panel data and the FD
estimator).

But even if $\Delta x_{i}$ is not constant, it might not vary much,
leading to large standard errors.

::: callout-note
### Practise Question

Consider the setup where you have a panel dataset for many individuals
$i$ and two time periods for each individual. All respondents are at
least 30 years old.

-   $y_{it}$ is log earnings (`lhrpay`)

You want to investigate the impact of a number of variables on $y$.
Which of the following are variables, whose relation to $y$ **cannot**
be investigated by a first difference estimation?

```{r}
#| echo: false
opts_p <- c(
   answer = "gender",
   answer = "age",
   "level of training",
   "distance between home and workplace"
)

```

`r longmcq(opts_p)`

`r hide("Can you help me understand the answer")` The key to understand
here is that the variable needs to have variation across all $i$
**after** differencing. Neither age nor gender have such variation.
Everyones gender will remain unchanged and everyone will be one year
older if observations are apart by a year. `r unhide()`
:::

Measurement error gets worse with differencing. We know from the
Endogenity section that means that the FD estimator gets more biased.

What this type of estimation basically does is to eliminate the variation in the explanatory variable $x_{it}$ between the different observations $i$. This is obvious if you were to think about the following three individuals.

|indiv | $x_{t=1}$| $x_{t=2}$|$\Delta x$|
|:------|:------|:------|:------|
|John  |15 | 17| 2|
|Xu | 4| 6 | 2 |
|Ludmilla | 3 | 7 | 4 |

There is a lot of variation between John and Xu, namely that John has a much higher average value of $x$, however, after calculating the difference there all we see is that they both have the same change $\Delta x$. So all the information we are using is the change "within" an individual. The "between" variation has been removed.

# Fixed-effects estimator (FE)

As we just discussed, differencing has removed the variation between individuals from the model. The effect size $\beta_1$ will be calculated using "within" variation only. This is why we do need that the values of $x_i$ actually change through time. Otherwise there would be no such variation.

There is an alternative way to achieve the same.

We could include a dummy variable for each individual (or for each minus 1) to avoid the dummy variable trap. We also say that we include a **fixed-effect** for individuals.

# Multi-period panels

The setup above is for a two-period panel. Things do not change
substantially when we have data for more than two periods.

| $i$    | $t$    | $y_{it}$ | $1$      | $d^1_t$ | $d^2_t$ | $d^3_t$ | $x_{it}$ | $\Delta y_{it}$ | $\Delta 1$ | $\Delta d^1_{t}$ | $\Delta d^2_{t }$ | $\Delta d^3_{t }$ | $\Delta x_{it}$ |
|--------|--------|----------|--------|-------------|-------------|-------------|----------|-----------------|------------|------------------|-------------------|-------------------|-----------------|
| 1      | 1      | $y_{11}$ | 1      | 1           | 0           | 0           | $x_{11}$ | .               | .          | .                | .                 | .                 | .               |
| 1      | 2      | $y_{12}$ | 1      | 0           | 1           | 0           | $x_{12}$ | $\Delta y_{12}$ | 0          | --1              | 1                 | 0                 | $\Delta x_{12}$ |
| 1      | 3      | $y_{13}$ | 1      | 0           | 0           | 1           | $x_{13}$ | $\Delta y_{13}$ | 0          | 0                | --1               | 1                 | $\Delta x_{13}$ |
| 2      | 1      | $y_{21}$ | 1      | 1           | 0           | 0           | $x_{21}$ | .               | .          | .                | .                 | .                 | .               |
| 2      | 2      | $y_{22}$ | 1      | 0           | 1           | 0           | $x_{22}$ | $\Delta y_{22}$ | 0          | --1              | 1                 | 0                 | $\Delta x_{22}$ |
| 2      | 3      | $y_{23}$ | 1      | 0           | 0           | 1           | $x_{23}$ | $\Delta y_{23}$ | 0          | 0                | --1               | 1                 | $\Delta x_{23}$ |
| ... | ... | ...   | ... | ...      | ...      | ...      | ...   | ...          | ...     | ...           | ...            | ...            | ...          |
| $N$    | 1      | $y_{N1}$ | 1      | 1           | 0           | 0           | $x_{N1}$ | .               | .          | .                | .                 | .                 | .               |
| $N$    | 2      | $y_{N2}$ | 1      | 0           | 1           | 0           | $x_{N2}$ | $\Delta y_{N2}$ | 0          | --1              | 1                 | 0                 | $\Delta x_{N2}$ |
| $N$    | 3      | $y_{N3}$ | 1      | 0           | 0           | 1           | $x_{N3}$ | $\Delta y_{N3}$ | 0          | 0                | --1               | 1                 | $\Delta x_{N3}$ |


There are now 3 time dummies, not 2. Time dummies are important as they control for macroeconomic effects, but, due to the dummy variable trap, we can only include two of these, implying that we estimate:

$$
y_{it} = \beta_0 + \delta_2 d^2_t + \delta_3 d^3_t + \beta_1 x_{it} + a_i + u_{it}
\qquad t=1,2,3
$$

When we estimate this model by FD, then this is the same as estimating

$$
\Delta y_{it} = \delta_2 \Delta d^2_t + \delta_3 \Delta d^3_t  + \beta_1 \Delta x_{it} + \Delta u_{it} \qquad t=2,3
$$

using POLS, pooling over the differenced observations we have for $t=2,3$. As you can see, all right hand side variables need to be differenced.





# Twins Double the Fun

Twinsburg, Ohio, near Cleveland, was founded as Millsville in the early
nineteenth century. Prosperous Millsville businessmen Moses and Aaron
Wilcox were identical twins whom few could distinguish. Moses and Aaron
were generous to Millsville in their success, a fact recognized when
Millsville was renamed Twinsburg in the early nineteenth century. Since
1976, Twinsburg has embraced its zygotic heritage in the form of a
summer festival celebrating twins. Millsville's annual Twins Days
attract not only twins reveling in their similarities but also
researchers looking for well-controlled comparisons.

Twin siblings indeed have much in common: most grow up in the same
family at the same time, while identical twins even share genes. Twins
might therefore be said to have the same ability as well. Perhaps the
fact that one twin gets more schooling than his or her twin sibling is
due mostly to the sort of serendipitous or random forces. The notion
that one twin provides a good control for the other motivates a pair of
studies by Ashenfelter and Krueger (1994) as well as Ashenfelter and
Rouse (1998). The key idea behind this work, as in many other studies
using twins, is that if ability is common to a pair of twin siblings, we
can remove it from the equation by subtracting one twin's data from the
other's and working only with the differences between them.

The long regression that motivates a twins analysis of the returns to
schooling can be written as

\$\$ \log y\_{fi} = \beta\_0 + \beta*1 s*{fi} + \mathbf{x}'*f*
\mathbf{b} + a_f + u{fi} \qquad i=1,2

\$\$

Here, subscript $f$ stands for family, while subscript $i = 1,2$ indexes
twin siblings, say, Ronald ($i=1$) and Donald ($i=2$). When Ronald and
Donald have the same ability, we can simplify by writing $a_{fi} = a_f$
and restating the above model seperately for both twins ($i=1$ and
$i=2$).

$$
\begin{align*}
\log y_{f1} &= \beta_0 + \beta_1 s_{f1} + \mathbf{x}'_f \mathbf{b} + a_f + u_{f1}\\
\log y_{f2} &= \beta_0 + \beta_1 s_{f2} + \mathbf{x}'_f \mathbf{b} + a_f + u_{f2}
\end{align*}
$$

Subtracting the equation for Donald from that for Ronald gives

$$
\begin{equation}
  \Delta \log y_{f} = \delta_0  + \beta_1 \Delta s_{f} + \Delta u_{f}.
\end{equation}
$$

an equation from which ability disappears
($\Delta \log y_{f} = \log y_{f2} - \log y_{f1}$ and
$\Delta s_{f}= s_{f2}-s_{f1}$). From this we learn that when ability is
constant within twin pairs, a short regression of the difference in
twins' earnings on the difference in their schooling recovers the long
regression coefficient, $\beta_1$.

Regression estimates constructed without differencing in the twins
sample generate a schooling return of about 11%, remarkably similar to
Mincer's estimate for the return to education. This can be seen in the
first column of Table 6.2 (@fig_twins). The model that produces the
estimates in column (1) includes age, age squared, a dummy for women,
and a dummy for whites. White twins earn less than black twins, an
unusual result in the realm of earnings comparisons by race, though the
gap here is not significantly different from zero.

The differenced equation (6.5) generates a schooling return of about 6%,
a result shown in column (2) of Table 6.2. This is substantially below
the short regression estimate in column (1). This decline may reflect
ability bias in the short model. Yet, once again, more subtle forces may
also be at work.

![Copy of Table 6.2 in Mastering
Metrics](../../images/MMtbl62.png){#fig-twins fig-alt="ALT"}

\begin{footnotesize}

Footnote~6.  Orley Ashenfelter and Alan B. Krueger, ``Estimates of the
Economic Returns to Schooling from a New Sample of Twins'', American
Economic Review, vol. 84, no. 5, December 1994, pages 1157--1173, and
Orley Ashenfelter and Cecilia Rouse, ``Income, Schooling, and Ability:
Evidence from a New Sample of Identical Twins'', Quarterly Journal of
Economics, vol. 113, no. 1, February 1998, pages 253--284.

Footnote~7. Estimates of this differenced model can also be obtained by adding a
dummy for each family to an undifferenced model fit in a sample that
includes both twins.  Family dummies are like selectivity-group dummies
in equation (2.2) in Chapter 2 and state dummies in equation (5.5) in
Section 5.2.  With only two observations per family, models estimated
after differencing across twins within families to produce a single
observation per family generate estimates of the returns to schooling
identical to those generated by ``dummying out'' each family in a pooled
sample that includes both twins.

MJA also notes that he has added a constant $\delta_0$ that isnt in MM
(6.5). This is because one \ul{always} needs a constant unless the sample
means are zero, which need not be the case here.

\end{footnotesize}

\subsubsection*{Twin Reports from Twinsburg}

Twins are similar in many ways, including---alas---their schooling. Of
340 twin pairs interviewed for the Twinsburg schooling studies, about
half report identical educational attainment. Schooling differences,
$\Delta s_{f}$, vary much less than schooling levels, $s_{if}$. If most
twins really have the same schooling, then a fair number of the nonzero
differences in reported schooling may reflect mistaken reports by at
least one of them. Erroneous reports, called \textbf{measurement
error}, tend to reduce estimates of $\beta_1$ in equation (6.5), a fact
that may account for the decline in the estimated returns to schooling
after differencing. A few people reporting their schooling incorrectly
sounds unimportant, yet the consequences of such measurement error can
be major.

To see why mistakes matter, imagine that twins from the same family
always have the same schooling. In this scenario, the only reason
$\Delta s_{f}$ isn't zero for everyone is because schooling is sometimes
misreported. Suppose such erroneous reports are due to random
forgetfulness or inattention rather than something systematic. The
coefficient from a regression of earnings differences on schooling
differences that are no more than random mistakes should be zero since
random mistakes are unrelated to wages. In an intermediate case, where
some but not all of the variation in observed schooling is due to
misreporting, the coefficient in equation (6.5) is smaller than it would
be if schooling were reported correctly. The bias generated by this sort
of measurement error in regressors is called \textbf{attenuation
bias}. The mathematical formula for attenuation bias is derived in the
chapter appendix. \[MJA insert: see Section\~\ref{sub:ME} above.\]

Misreported schooling attenuates the levels regression estimates shown
in column (1) of Table 6.2, but less so than the differenced estimates
in column (2). This difference in the extent of attenuation bias is also
illustrated by the hypothetical scenario where all twins share the same
schooling but schooling levels differ across families. When twins in the
same family really have the same schooling, all variation in
within-family differences in reported schooling comes from mistakes. By
contrast, most of the cross-family variation in reported schooling
reflects real differences in education. Real variation in schooling is
related to earnings, a fact that moderates attenuation bias in estimates
of the model for levels, equation (6.4). This reflects a general point
about the consequences of covariates for models with mismeasured
regressors---additional controls make attenuation bias worse---a point
detailed in the chapter appendix.

Measurement error raises an important challenge for the Twinsburg
analysis, since measurement error alone may explain the pattern of
results seen in columns (1) and (2) of Table 6.2. Moving from the levels
to the differenced regression accentuates attenuation bias, probably
more than a little. The decline in schooling coefficients across columns
may therefore have little to do with ability bias. Fortunately, seasoned
masters Ashenfelter, Krueger, and Rouse anticipated the attenuation
problem. They asked each twin to report not only their own schooling but
also that of their sibling. As a result, the Twinsburg data sets contain
two measures of schooling for each twin, one self-report and one sibling
report. The sibling reports provide leverage to reduce, and perhaps even
eliminate, attenuation bias.

The key tool in this case, as with many of the other problems we've
encountered, is IV. \[MJA insert: We pick this up again in
Section\~\ref{sub:IV}.\]

# Reading

The material for the three biases comes from Wooldridge (2025). See the
subsections on Omitted variable bias, Properties of OLS under
Measurement Error, and the section Simultaneity Bias in OLS.

Angrist, J.D. and J.-S. Pischke (2015) Mastering Metrics. Princeton
University Press.

Ashenfelter, O. and A. Krueger (1994) Estimates of the Economic Return
to Schooling from a New Sample of Twins, The American economic review,
84(5), 1157 - 1173.

Ashenfelter, O. and C. Rouse, (1998) Income, Schooling, and Ability:
Evidence from a New Sample of Identical Twins, The Quarterly journal of
economics, 113(1), 253 - 284.

Cunningham, S. (2021) [Causal Inference: The
MixTape](https://mixtape.scunning.com/index.html)

Wooldridge, J. (2025) Introductory Econometrics: A Modern Approach,
Cengage.
