---
title: "Panel Data Methods"
format: 
  html:
    toc : true
    css: [webex.css]
    include-after-body: [webex.js]
    embed-resources: true 
editor: 
  render-on-save: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(webexercises)
```

```{r, echo = FALSE, results='asis'}
# Uncomment to change widget colours:
#style_widgets(incorrect = "goldenrod", correct = "purple", highlight = "firebrick")
```

# Introduction to Panel Data Structures

We shall begin by introducing Panel Data structures. Being clear about data structures is a prerequisite for successful empirical work.

Previously we discussed that parameter estimator bias can be caused by omitted variables sometimes also called unobserved heterogeneity. There will be situations when using panel data will give us a route to removing this bias. 

## Two-period panel data

Suppose we have a dataset comprising $N$ individuals (or other micro unit) observed exactly $T=2$ times.

| $i$    | $t$    | $y_{it}$ | $1$      | $d^1_t$ | $d^2_t$ | $x_{it}$ | $a_{i}$ | $u_{it}$ |
|--------|--------|----------|--------|---------|---------|----------|---------|----------|
| 1      | 1      | $y_{11}$ | 1      | 1       | 0       | $x_{11}$ | $a_{1}$ | $u_{11}$ |
| 1      | 2      | $y_{12}$ | 1      | 0       | 1       | $x_{12}$ | $a_{1}$ | $u_{12}$ |
| 2      | 1      | $y_{21}$ | 1      | 1       | 0       | $x_{21}$ | $a_{2}$ | $u_{21}$ |
| 2      | 2      | $y_{22}$ | 1      | 0       | 1       | $x_{22}$ | $a_{2}$ | $u_{22}$ |
| ... | ... | ...   | ... | ...  | ...  | ...   | ...  | ...   |
| $N$    | 1      | $y_{N1}$ | 1      | 1       | 0       | $x_{N1}$ | $a_{N}$ | $u_{N1}$ |
| $N$    | 2      | $y_{N2}$ | 1      | 0       | 1       | $x_{N2}$ | $a_{N}$ | $u_{N2}$ |


The variables we observe are

* $y_{it}$ and $x_{it}$; both vary over both $i$ and $t$.
* $d^1_t$ is a period 1 time dummy; it only varies over $t$.
* $d^2_t$ is a period~2 time dummy; it only varies over $t$.
* $1$ is the constant.

whereas the variables not observed are:

* $a_i$ does not vary over $t$. It is a so-called **fixed effect**, or **unobserved heterogeneity**. Think about a personal characteristic of subject $i$ that is constant through time. This could be things like general levels of ability or intelligence.
* $u_{it}$ (not shown) is an **idiosyncratic error**. It is uncorrelated with everything (like most error terms). In particular:

$$
Cov(x_{i1},u_{i1})=Cov(x_{i2},u_{i2})=0 ~~~(AFD1)
$$ 

The crucial problem is that the omitted variable $a_i$ is correlated with $x_{it}$. For example, in the wage-education model in
the Omitted Variable Bias Section (now with slightly different notation but essentially the same):

$$
y_{it} = \beta_0 + \delta_0 d^2_t + \beta_1 x_{it} + \beta_2 a_{i} + u_{it}.
$$ {#eq-woods}

where

* $y_{it}$ is log earnings (`lhrpay`)
* $x_{it}$ is years of schooling (`educ`)
* $a_{i}$ is unobserved innate ability (`abil`)

and we also included a dummy variable for the 2nd period, $d^2_t $. Because of the dummy-variable trap, one of two time dummies must be dropped, here $d^1_t$.  (Recall one cannot have a constant, male dummy __and__ a female dummy in the same regression.)


Here, ability is correlated with education. If $a_i$ were __not__ correlated with $x_{it}$, the OVB formula says OLS in unbiased
and we wouldn't have a problem in the first place. The way that panel data can help is through the crucial assumption that $a_i$ is constant over time. This is an untestable assumption as $a_i$ is not observed.

The above equation is also called a **fixed effects** model.


# The First Difference estimator (FD) {#sec-FD}

The First Difference estimator makes use of this setup to estimate $\beta_1$.

## Differencing out $a_i$

We could estimate @eq-woods by OLS, treating $a_i + u_{it}$ as a single **composite error**.  This estimator is called **Pooled OLS** (POLS).  However, we have already assumed that $Cov(x_{it},a_i)\neq 0$ which means that Pooled OLS is biased (and inconsistent).  This is often called \textbf{heterogeneity bias}.

__However__, with panel data, we can difference-out the unobserved fixed effect $a_i$. Let's write @eq-woods for both time periods $t=1$ and $t=2$.

$$
\begin{gather*}
y_{i1} = \beta_0            + \beta_1 x_{i1}  + a_i + u_{i1} \quad (t=1)\\
y_{i2} = \beta_0 + \delta_0 + \beta_1 x_{i2}  + a_i + u_{i2} \quad (t=2)
\end{gather*}
$$

Subtracting the $t=1$ equation from the $t=2$ equation gives

$$
\begin{equation}
  \Delta y_{i} = \delta_0  + \beta_1 \Delta x_{i} + \Delta u_{i}.
\end{equation}
$$ {#eq-mee}

To see what is happening, look at the schematic data above, and subtract the first row from the second for each $i$:

| $i$    | $t$    | $y_{it}$ | 1      | $d^1_{t}$   | $d^2_{t}$   | $x_{it}$ | $\Delta y_{it}$ | $\Delta 1$ | $\Delta d^1_{t}$   | $\Delta d^2_{t}$   | $\Delta x_{it}$ |
|--------|--------|----------|--------|-------------|-------------|----------|-----------------|------------|--------------------|--------------------|-----------------|
| 1      | 1      | $y_{11}$ | 1      | 1           | 0           | $x_{11}$ | .               | .          | .                  | .                  | .               |
| 1      | 2      | $y_{12}$ | 1      | 0           | 1           | $x_{12}$ | $\Delta y_{1}$  | 0          | --1                | 1                  | $\Delta x_{1}$  |
| 2      | 1      | $y_{21}$ | 1      | 1           | 0           | $x_{21}$ | .               | .          | .                  | . .                | .               |
| 2      | 2      | $y_{22}$ | 1      | 0           | 1           | $x_{22}$ | $\Delta y_{2}$  | 0          | --1                | 1                  | $\Delta x_{2}$  |
| ... | ... | ...   | ... | ...      | ...      | ...   | ...          | ...     | ...             | ...             | ...          |
| $N$    | 1      | $y_{N1}$ | 1      | 1           | 0           | $x_{N1}$ | .               | .          | .                  | .                  | .               |
| $N$    | 2      | $y_{N2}$ | 1      | 0           | 1           | $x_{N2}$ | $\Delta y_{N}$  | 0          | --1                | 1                  | $\Delta x_{N}$  |

Expressing @eq-mee in words, the change in $y_{it}$ between periods 1 and 2 is regressed on the change in the regressor(s) and a constant, using $N$ obs. Note that
the variables are not defined for period 1 and so the dataset comprises just $N$ obs, whereas regressing @eq-woods could be done on $2N$ observations. 

Just to be clear about the variables that are being used in @eq-mee:

* $\Delta y_{i} \equiv y_{i2}-y_{i1}$ is the \textbf{difference} of $y_{i}$.
* Ditto $\Delta x_{i} \equiv x_{i2}-x_{i1}$.  (In a multiple regression all the regressors are differenced.)
* $\beta_0$ and $\beta_0+\delta_0$ are "macro" effects in that they affect all individuals identically, in the same time period.  But only $\delta_0$ is identified.  $\beta_0$  drops out because $\Delta 1 =0 = \Delta d^1_{t} +\Delta d^2_{t}$.
* $\delta_0$ is the parameter on $\Delta d^2_t$, which is the new regression constant.

When OLS is used on this transformed model, it is labelled the **first-difference estimator** (FD).  In this example, it is a Simple Regression. In general, FD is a multiple regression where **all** variables are differenced.

The important issue is whether the assumptions for OLS are satisfied?  If yes, FD is unbiased.  Recall that the assumptions are:

* The RHS variable is exogenous (uncorrelated with the error);
* The RHS variable cannot be constant.

## Exogeneity

The first assumption asks whether $\Delta x_{i}$ is uncorrelated with $\Delta u_{i}$?

$$
\begin{align*}
Cov (\Delta x_i, \Delta u_i) &=
Cov [(x_{i2}-x_{i1}),(u_{i2}-u_{i1})] \\
& =Cov(x_{i2},u_{i2}) +
Cov(x_{i1},u_{i1}) -
Cov(x_{i1},u_{i2}) -
Cov(x_{i2},u_{i1}) \\ 
&=0.
\end{align*}
$$

All 4 covariances need to be zero.  We have already assumed the first pair are zero (AFD1).  If AFD1 is true, there is no reason why the second pair ("across" $t=1,2$) should not be zero either.

To summarise, if $x$ and $u$ are uncorrelated, then so are $\Delta x$ and $\Delta u$. But this is importantly an assumption. One needs to assume that there is no unobserved variation in the error terms that is correlated to the $x$ variable. What has been gained through the first-difference setup is that any such variation in the error term that is constant through time by $i$ has been removed. We do not have to worry any longer about such unobserved variation.

::: {.callout-note}
### Practise Question

Consider the setup where

* $y_{it}$ is log earnings (`lhrpay`)
* $x_{it}$ is a variable encoding how many hours individual $i$ spent on learning new skills in period $t$
Which of the following potentially unobserved heterogeneity will be removed/taken care of by applying a FD estimation in a context where you have a panel dataset for many individuals $i$ and two time periods for each individual.

```{r}
#| echo: false
opts_p <- c(
   answer = "different general ability by individual $i$",
   "different levels of economic growth in individual $i$'s area",
   answer = "different personalities of individuals $i$",
   "different life circumstances for individuals $i$"
)

```

`r longmcq(opts_p)`

`r hide("Can you help me understand the answer")`
It is important to understand that only unobserved variation that remains constant for individual $i$ across the time periods is differenced out. This is why economic growth (which varies through time) and life circumstances which often change from one year to the next, are not controlled for by using a FD estimation.
`r unhide()`

:::

## Variation in $\Delta x_i$

The second assumption says that each $\Delta x_{i}$ must have some variation across $i$.  Age, gender, and education are good examples of
$\Delta x_{i}$ being constant.  To see what is happening, observe what happens when we first-difference the male dummy $m$, age $a$ and
education $e$:

| $i$    | $t$    | $y_{it}$ | $d^2_t$ | $m$    | $a$    | $e$    |   | $\Delta y_{it}$ | $\Delta d^2_{t}$ | $\Delta m$ | $\Delta a$ | $\Delta e$? |
|--------|--------|----------|---------|--------|--------|--------|---|-----------------|--------------------|------------|------------|-------------|
| 1      | 1      | $y_{11}$ | 0       | 1      | 18     | 11     |   | .               | .                  | .          | .          |
| 1      | 2      | $y_{12}$ | 1       | 1      | 24     | 11     |   | $\Delta y_{1}$  | 1                  | 0          | 6          |
| 2      | 1      | $y_{21}$ | 0       | 0      | 28     | 13     |   | .               | . .                | .          | .          |
| 2      | 2      | $y_{22}$ | 1       | 0      | 34     | 13     |   | $\Delta y_{2}$  | 1                  | 0          | 6          |
| ... | ... | ...   | ...  | ... | ... | ... |   | ...          | ...             | ...     | ...     |
| $N$    | 1      | $y_{N1}$ | 0       | 1      | 51     | 16     |   | .               | .                  | .          | .          |
| $N$    | 2      | $y_{N2}$ | 1       | 1      | 57     | 16     |   | $\Delta y_{N}$  | 1                  | 0          | 6          |


This is important, because some issues (education, discrimination etc) cannot be investigated using this approach (panel data and the FD
estimator).

But even if $\Delta x_{i}$ is not constant, it might not vary much, leading to large standard errors.

::: {.callout-note}
### Practise Question

Consider the setup where you have a panel dataset for many individuals $i$ and two time periods for each individual. All respondents are at least 30 years old.

* $y_{it}$ is log earnings (`lhrpay`)

You want to investigate the impact of a number of variables on $y$. Which of the following are variables, whose relation to $y$ **cannot** be investigated by a first difference estimation?

```{r}
#| echo: false
opts_p <- c(
   answer = "gender",
   answer = "age",
   "level of training",
   "distance between home and workplace"
)

```

`r longmcq(opts_p)`

`r hide("Can you help me understand the answer")`
The key to understand here is that the variable needs to have variation across all $i$ **after** differencing. Neither age nor gender have such variation. Everyones gender will remain unchanged and everyone will be one year older if observations are apart by a year. 
`r unhide()`


:::

Measurement error gets worse with differencing. We know from the Endogenity section that means that the FD estimator gets more biased.


# Reading

The material for the three biases comes from Wooldridge (2025). See the subsections on Omitted variable bias, Properties of OLS under
Measurement Error, and the section Simultaneity Bias in OLS.

Angrist, J.D. and J.-S. Pischke (2015) Mastering Metrics. Princeton University Press.

Cunningham, S. (2021) [Causal Inference: The MixTape](https://mixtape.scunning.com/index.html)

Wooldridge, J. (2025) Introductory Econometrics: A Modern Approach, Cengage.
